---
title: "Learning knitr"
author: "andersc"
date: "Tuesday, August 12, 2014"
output: html_document
---

## Problems

This is no single document to integrate data analysis with textual representations; i.e. data, code, and text are not linked.

## knitr

*Literate programs* are **weaved** to produce human-readable documents and **tangled** to produce machine-readable documents.

**Literate programming** is a general concept, we need

* a documentation language
* a programming language

The original **Sweave** system develped by Frienrich Leisch used LaTeX and R.

## Pros of Literate Programming

* Text and code all in one place, logical order

## Cons of Literate Programming

* If there are too much code in the document
* Slow down processing of docs

## knitr package

* Supports RM, LaTeX, and HTML
* Can export to PDF, HTML
* Built right into RStudio

## Good For

* Manuals
* Short/medium length technical docs
* Tutorials
* Reports
* Data preprocessing docs

## Not Good For

* Very long research articles
* Complex time-consuming computations
* Docs that require precise formatting



## Show R codes

```{r, results='asis'}
set.seed(1)
x <- rnorm(100)
mean(x)
```

## Manually Export

```{r, echo=FALSE, results='hide'}
#library(knitr)
#setwd("")
#knit2html("learning.Rmd")
#browseURL("learning.html")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


